\documentclass{amsart}

\usepackage{fouche}
\input{macros.tex}

\newenvironment{fo}[1]{\color{red} \textbf{Fouche said}: #1}{\color{black}}
\newenvironment{de}[1]{\color{blue} \textbf{Denta said}: #1}{\color{black}}

\newcommand{\topicRule}{\todo[cyan, inline]{}}
\usepackage{booktabs}
\begin{document}
\begin{abstract}
  The present paper is the first piece of a series whose aim is to develop an approach to ontology and metaontology through category theory. More in detail, we exploit the theory of \emph{elementary toposes} to purport the claim that a satisfying ``theory of existence'', and more at large ontology itself, can be obtained appealing to category theory. In this perspective, ontology possesses no unitary nature; it is instead a category, the universe of discourse in which our mathematics (intended at large as a theory of knowledge) can be deployed. The \emph{internal language} that all categories possess prescribes the theory of existence of a fixed ontology/category.

  In such internal language of a topos, all first-order properties of an object can be modeled by propositions $p : U \to\Omega$ valued in an object whose ``elements'' are the truth values that $p$ can assume. Each ontology $\clE$ yields its own $\Omega_\clE$, and thus its own set of truth values, and thus its own ``theory of meaning''. This approach resembles, but is more general than, fuzzy logics, as suitable choices of $\clE$ and thus of $\Omega_\clE$ carry more nuanced truth values than the usual ``true'' or ``false'' of classical logic.

  Similarly to the mathematical approach used in the solution of physical problems, a solid corpus of structural mathematics can be used to yield nontrivial philosophical consequences and a useful, modular language suitable for a practical foundation framing some essential, elementary questions of ontology in their correct position. As both a test-bench for our theory, and a literary \emph{divertissement}, we propose a possible category-theoretic solution of Borges' famous paradoxes of Tl\"on's ``nine copper coins'', and of other seemingly paradoxical construction in his literary work.
\end{abstract}
\maketitle
  \tableofcontents
  \input{sections/01-introduction.tex}
  \input{sections/02-variable-sets.tex}
  \input{sections/03-internal-language.tex}
  \input{sections/04-nine-coins.tex}
  \input{sections/05-vistas.tex}
  \appendix
  \input{sections/AA-categories.tex}

  \bibliography{allofthem}{}
  \bibliographystyle{amsalpha}
\end{document}
