\section{Nine copper coins, and other toposes}\label{sec_coins}
\epigraph{
	Explicaron que una cosa es \emph{igualdad}, y otra \emph{identidad}, y formularon una especie de \emph{reductio ad absurdum}, o sea el caso hipotético de nueve hombres que en nueve sucesivas noches padecen un vivo dolor. ¿No sería ridículo -interrogaron- pretender que ese dolor es el mismo?
}{JLB ---Tl\"on, Uqbar, Orbis Tertius}

According to our description of the Mitchell-Bénabou language in the category of variable sets, \emph{propositions} are morpisms of the form
\[p : U \to \Omega_I\]

where $\Omega_I$ is the subobject classifier of $\Set/I$ described in \autoref{variable_sets_have_omega}; now, recall that
\begin{itemize}
	\item the object $\Omega_I = \{0,1\}\times I \to I$ becomes an object of $\Set/I$ when endowed with the projection $\pi_I : \Omega_I \to I$ on the second factor of its domain;
	\item the universal monic $\tr : I \to \Omega_I$ consists of a section of $\pi_I$, precisely the one that sends $i : I$ to the pair $(i,1) : \Omega_I$;
	\item every subobject $U \hookrightarrow A$ of an object $A$ results as a pullback (in $\Set/I$) along $\tr$:
	      \[\xymatrix@R=5mm@C=5mm{
		      U\ar[dr]^u\ar[rr]^u\ar[dd]_m&& I\ar[dd]^{\tr} \ar@{=}[dl]\\
		      &I& \\
		      A \ar[ur]\ar[rr]_{\chi_U}&& \ar[ul]\Omega_I
		      }\]
	      (see \autoref{variable_sets_have_omega} for a complete proof)
\end{itemize}
The set $I$ in this context acts as a \emph{multiplier} of truth values, in that every proposition can have a pair $(\epsilon, i)$ as truth value. We introduce the following notation: a proposition $p : U \to \Omega_I$ is \emph{true}, in context $x :U$, with \emph{strength} $t$, if $p(x) =(1,t)$ (resp., $p(x)=(0,t)$).
\begin{remark}\label{very_importanta_force}
A proposition in the internal language of variable sets is a morphism of the following kind: a function $p : U \to \Omega_I$, defined on a certain domain, and such that
\[
	\xymatrix{
		U\ar[d]_u\ar[r]^-p  & \{0,1\}\times I \ar[d]^{\pi_I}\\
		I \ar@{=}[r]& I
	}
\]
(it must be a morphism of variable sets!) This means that $\pi p(x : U) = u(x : U)$, so that $p(x) = (\epsilon_x, u(x))$ for $\epsilon_x =0,1$ and $u$ is uniquely determined by the "variable domain" $U$. This is an important observation: the strength with which $p$ is true/false is completely determined by the structure of its domain, in the form of the function $u : U \to I$ that renders the pair $(U,u)$ an object of $\Set/I$.
\end{remark}
To get a grip of the different roles of various classes of propositions, and given that our interest will be limited to a certain class of particular propositions that we will construct \emph{ex nihilo}, it is now convenient to discuss what constraints we have to put on the structure of $I$: of course, the richest this structure is, the better will the category $\Set/I$ behave: it is for example possible to equip $I$ with an order structure, or a natural topology. Among different choices of truth multiplier, yielding different categories of variable sets, and different kinds of internal logic therein, we will privilege those that make $I$ behave like a space of strengths: a dense, linear order with LUP, thus not really far from being a closed, bounded subset of the real line.

The main result of the present section is a roundup of examples showing that it is possible to concoct categories of variable sets where some seemingly paradoxical constructions coming from J.L. Borges' literary world have, instead, a perfectly ``classical'' behaviour when looked with the lenses of the logic of variable set theory.

Each of the examples in our roundup \autoref{bla} \autoref{bli} \autoref{blu} is organised as follows: we recall the shape of a paradoxical statement in Borges' literary world. Then, we show in which topos this reduces to an intuitive statement expressed in the syntax of a variable set category.

More than often, we use $I=[0,1]$ as base of variable sets; as already said, there are different reasons for this choice: the most intuitive is that if a truth value is given with a \emph{strength} $t\in I$ it is a natural request to be able to \emph{compare} elements in this set; in particular, it should always be possible to assess what truth value is stronger.

For this reason, even if this assumption is never strictly necessary (the only constraint is that $I$ is totally, or partially, ordered set by a relation $\le$), a natural choice for $I$ is a \emph{continuum} (=a dense total order with LUP --see \cite{moschovakis2009descriptive}). An alternative choice drops the density assumption: in that case the (unique) finite total order $\Delta[n] = \{0 < 1 <\dots < n\}$, or the countable total order $I=\omega = \bigcup_n \Delta[n]$ are all pretty natural choices for $I$ (although it is way more natural for $I$ to have a minimum \emph{and} a maximum element).\footnote{We're oly interested in the notion of an abstract interval here: a continuum $X$ endowed with an operation $X \to X \vee X$ of ``zooming'', uniquely defined by this property. In a famous paper Freyd characterises ``the interval'' as the terminal interval coalgebra: see \cite[§1]{freyd2008algebraic}; for our purposes, note that $[0,1]$ is a natural choice: it is a frame, thus a Heyting algebra $\fkH =([0,1],\land,\lor,\Rightarrow)$ with respect to the pseudo-complement operation given by $(x \Rightarrow z) := \bigvee_{x\land y \le z} y$ (it is immediate that $x \land a \le b$ if and only if $a \le x \Rightarrow b$ for every $a,b\in [0,1]$).} In each of these cases ``classical'' logic is recovered as a projection: propositions $p$ can be true or false with strength $1$,\footnote{Here $I$ is represented as an interval whose minimal and maximal element are respectively $0$ and $1$; of course these are just placeholders, but it is harmless for the reader to visualise $I$ as the interval $[0,1]$.} the maximum element of $I$:
\begin{center}
	\begin{tikzpicture}
		\draw[gray!70,dashed] (2,1.25) -- (2,-.25) node[below] {$\{\perp,\top\}$};
		\draw[fill] (0,0) circle (1pt) node[left] {$0$};
		\draw[fill] (2,0) circle (1pt) node[right] (dis) {$1$};
		\draw (0,0) -- (2,0);
		\begin{scope}[yshift=1cm]
			\draw[fill] (0,0) circle (1pt) node[left] {$0$};
			\draw[fill] (2,0) circle (1pt) node[right] (dat) {$1$};
			\draw (0,0) -- (2,0);
			\node[right of=dis] {$\{\top\}\times I$};
			\node[right of=dat] {$\{\perp\}\times I$};
		\end{scope}
	\end{tikzpicture}
\end{center}
In order to aid the reader understand the explicit way in which $I$ ``multiplies'' truth values, we spell out explicitly the structure of the subobject classifier in $\Set/\Delta[2]$. In order to keep calling the minimum and maximum of $I$ respectively $0$ and $1$ we call $\frac{1}{2}$ the intermediate point of $\Delta[2]$.
\begin{remark}
	The subobject classifier of $\Set/\Delta[2]$ consists of the partially ordered set $\Delta[1]\times\Delta[2]$ that we can represent pictorially as a rectangle
	\[\begin{tikzpicture}[xscale=2]
			\foreach \i/\name in {0/0,.5/{\frac{1}{2}},1/0}{
			\foreach \j/\pos in {0/below,1/above}
			\fill (\i,\j) ellipse (1pt and 2pt) node[\pos] {\tiny $(\name,\j)$};
			}
			\draw (0,0) rectangle (1,1);
			\draw (.5,1) -- (.5,0);
		\end{tikzpicture}\]
	endowed with the product order. The resulting poset is partially ordered, and in fact a Heyting algebra, because it results as the product of two Heyting algebras: the Boole algebra $\{0<1\}$ and the frame of open subsets of the Sierpinski space $\{a,b\}$ (the topology is $\tau_S = \{\varnothing, \{a\}, \{a,b\}\}$).
\end{remark}
\begin{remark}
	Siccome il caso $I=[0,1]$ con la topologia euclidea è quello piu naturale per diversi motivi, definiamo alcuni insiemi di interesse per una data proposizione $p : U \to \Omega_I$ per questa scelta di $I$:
	\begin{itemize}
		\item $A^\top = \{x : U \mid p(x) = (1,t_x), t_x > 0\} = p^\leftarrow(\{1\}\times (0,1])$ e $A^\perp = \{x : U \mid p(x) = (0,t_x), t_x > 0\}$; cose vere (risp., false) con forza maggiore di zero. Sono le funzioni $u : U \to I$ tali che $u^\leftarrow 0 = \varnothing$.
		\item $B^\top = \{x : U \mid p(x) = (1,1)\} = p^\leftarrow((1,1))$ e $B^\perp = \{x : U \mid p(x) = (0,1)\}$ cose vere (risp.,false).
		\item $E_t^\top = \{ x : U \mid p(x)=(1,t)\}$ e $E_t^\perp = \{ x : U \mid p(x)=(0,t)\}$; cose vere (risp., false) con forza $t$.
	\end{itemize}
\end{remark}
Last but not least, a crucial assumption will be that the strength of $p$ depends continuously, or not, on the variables on its domain of definition. Without such continuous dependence, small changes in context $x : U$ might drastically change the truth value $p(x)$.\footnote{There is no a priori reason to maintain that $p$ is a continuous proposition; one might argue that discontinuous changes in truth value of $p$ happen all the time in ``real life''; see the family of paradoxes based on so-called \emph{separating instants}: how well-defined the notion of ``time of death'' is? How well-defined the notion of ``instant in time''?}
\subsection{The unimaginable topos theory hidden in Borges' library}
Jorge Luis Borges' literary work is well-known for being made by paradoxical worlds; oftentimes, seemingly absurd consequences follow by stretching to their limit ideas from logic and mathematics: time, infinite, self-referentiality, duplication, recursion, the relativity of time, the illusory nature of our perceptions, the limits of language, its capacity to generate worlds.

In the present section we choose \emph{Fictions}, Borges' famous collection of novels, as source of inspiration for possible and impossible worlds and their ontology.

Usualmente la costruzione di un ``mondo impossibile'' va circa come segue: \dots ; noi rovesciamo tale prospettiva, e invece di depennare dal computo degli universi i mondi le cui caratteristiche do not comply with sensorial experience, or imply paradoxical entities/constructs, we accept their existence for bizarre that it may seem, and we try to deduce what kind of logic can consistently generate such statements.

The interest in such a literary calembour is manifold, and the results are surprising:
\begin{itemize}
	\item we unravel how a mathematically deep universe Borges has inadvertently created: of the many compromises we had to take in order to reconcile literature and the uniderlying mathematics,\footnote{See \autoref{our_beloved_interval} below: these ``compromises'' mainly amount to assumptions on the behaviour of space-time on Tl\"on and Babylon.} we believe no one is particularly far-fetched one;
	\item we unravel how \emph{relative} ontological assumptions are; they are not given: using category theory, ontology, far from being the presupposition on which it is based, is a byproduct of language itself. The more expressive language is, the more ontology; the fuzzier its capacity to assert truth, the fuzzier existence becomes;
	\item ``Fuzziness'' of existence, i.e. the fact \emph{entia} exist less than completely, is hard-coded in the language (in the sense of \autoref{da_lang}) of the category we decide to work in from time to time;
	\item ...
\end{itemize}
To sum up, readers willing to find an original result in this paper, might find it precisely here: we underline how Borges' alternative worlds (Babylon, Tl\"on \ dots) are mathematically consistent places, worthy of existence as much as our world, just based on a different internal logic. And they are so just thanks to a base-sensitive theory of existence --ontology breaks in a spectrum of ontolog\emph{ies}, one for each category/world.

The first paradox we aim to frame in the right topos is the famous nine copper coins argument, used by the philosophers of Tlön to construct a paradoxical object whose existence persists over time, in absence of a consciousness continually perceiving it and maintaining in a state of being.
\begin{example}[Nine copper coins]\label{bla}
	First, we recall the exact statement of the paradox from \cite{}:\footnote{The translation we employ is classical and comes from \cite{tlonEN}:
		\begin{quote}
			\hspace{.5em} Tuesday, $X$ crosses a deserted road and loses nine copper coins. On Thursday, $Y$ finds in the road four coins, somewhat rusted by Wednesday's rain. On Friday, $Z$ discovers three coins in the road. On Friday morning, $X$ finds two coins in the corridor of his house. The heresiarch would deduce from this story the reality - i.e., the continuity - of the nine coins which were recovered.

			\hspace{.5em} It is absurd (he affirmed) to imagine that four of the coins have not existed between Tuesday and Thursday, three between Tuesday and Friday afternoon, two between Tuesday and Friday morning. It is logical to think that they have existed - at least in some secret way, hidden from the comprehension of men - at every moment of those three periods.
		\end{quote}}
	\begin{quote}
		El martes, $X$ atraviesa un camino desierto y pierde nueve monedas de cobre.
		El jueves, $Y$ encuentra en el camino cuatro monedas, algo herrumbradas por la lluvia del miércoles. El viernes, $Z$ descubre tres monedas en el camino. El viernes de mañana, $X$ encuentra dos monedas en el corredor de su casa. El  quería deducir de esa historia la realidad -id est la continuidad- de las nueve monedas recuperadas.

		Es absurdo (afirmaba) imaginar que  cuatro de las monedas no han existido entre el martes y el jueves, tres entre el martes y la tarde del viernes, dos entre el martes y la madrugada del viernes. Es lógico pensar que han existido -siquiera de algún modo secreto, de comprensión vedada a los hombres- en todos los momentos de esos tres plazos.
	\end{quote}
	Before going on with our analysis, two remarks are in order:
	\begin{itemize}
		\item the paradox appears in a primitive version in \cite{}, where instead of nine copper coins, a single arrow, shot by an anonymous archer, disappears among the woods. The text appears in a hard-to-find edition of \emph{Inquisiciones} \cite{}; in the last chapter, we read the ``arrow avatar'' (\emph{avatar de la flecha}):
		      \begin{quote}
			      $X$ scocca una freccia da un arco, ed essa si perde fra gli alberi.

			      $X$ la cerca e riesce a ritrovarla.

			      E' assurdo immaginare che la freccia non sia esistita durante il periodo fra i momenti in cui $X$ l'ha persa di vista e l'ha ritrovata.

			      E' logico pensare che essa sia esistita - anche se in un certo modo segreto, di comprensione vietata agli uomini - in tutti i momenti di questo periodo.
		      \end{quote}
		\item There is one and only one reason why the paradox of the nine copper coins is invalid: copper does not rust.
	\end{itemize}
	It is obvious that both constructions leverage on the same argument to build an efficient aporia: the mysterious persistence of things through time without a perceivent consciousness. We concentrate on the copper coins dilemma, per il semplice fatto che Finzioni è raggiungibile molto piu facilmente ai nostri lettori; incidentally, we happen to be able to rectify the ``rust counterargument'' without appealing to the assumption that copper can rust on Tl\"on due to a difference in Tl\"onian chemistry.

	Expressed in natural language, our solution to the paradox goes more or less as follows: $X$ loses their coins on Tuesday, and the force $\varphi$ with which they ``exist'' lowers; it grows back in the following days, going back to a maximum value when $X$ retrieves two of their coins on the front door. $Y$ findings of other coins raises their existence force to a maximum. The coins that $Y$ has found rusted (or rather, the surface copper oxidized: this is possible, but water is rarely sufficient to ignite the process alone --certainly not in the space of a few hours.\footnote{in una sorta di principio di wormhole, eventi indipendenti sulla terra sono dipendenti su Tlön, perché l'evento A influenza, in uno spazio pluridimensionale di scelte di z di verità, l'evento B in modi che gli sarebbero vietati se fosse sulla Terra.}
\begin{remark}\label{our_beloved_interval}
	In this perspective, Tl\"on classifier of truth values can be taken as $\Omega_I = \{0<1\}\times I$, where $I$ is any set with more than one element; a minimal example can be $I=\{N,S\}$ (justifying this choice from inside Tl\"on is easy: the planet is subdivided into two emispheres; each of which now has its own logic ``line'' independent from the other), but as explained in \autoref{our_beloved_interval} a more natural choice for our purposes is the closed real interval $I=[0,1]$.

	This allows for a continuum of possible forces with which a truth value can be true or false;  it is tobe noted that $[0,1]$ is also the most natural place on which to interpret fuzzy logic, albeit the interest for $[0,1]$ therein can be easily and better motivated starting from probability theory. (But see \cite{} for an interesting perspective on how to develop basic measure teory out of $[0,1]$.)
\end{remark}
	We now start to formalise properly what we said until now.

	To set our basic assumptions straight, we proceed as follows:
	\begin{itemize}
		\item senza perdita di generalità possiamo supporre l'insieme $C = \{c_1,\dots,c_9\}$ delle monete totalmente ordinato e partizionato in modo tale che le prime due monete siano quelle ritrovate da $X$ il martedì, le seconde quattro quelle che $Y$ ritrova sul cammino, e le ultime 3 quelle viste da $Z$. Allora
		      \[C = C_X \sqcup C_Y \sqcup C_Z.\]
		      and $C_X = \{c_{X1}, c_{X2}\}$, $C_Y = \{
			      c_{Y1},c_{Y2},c_{Y3},c_{Y4}\}$, $C_Z= \{c_{Z1}, c_{Z2}, c_{Z3}\}$ As already said, the truth multiplier $I$ is the closed interval $[0,1]$ with its canonical order --so with its canonical structure of Heyting algebra, and if needed, endowed with the usual topology inherited by the real line.
		\item Propositions of interest for us are of the following form:
		      \[\lambda gcd.p(g, c, d) : \{X,Y,Z\}\times C\times W \to \Omega_I\]
		      where $W$ is a set of days, that for the sake of explicitness can be taken equal to the set of weekdays $S,M,T$ (strictly speaking, the paradox involves just the interval $[Tu,Fri]$). $p(g,c,d)$ has to be read as ``in $g$'s frame of existence the coin $c$ exists at day $d$ with strength $p(g,c,d)$''
	\end{itemize}
	Definiamo ora \emph{ammissibile} una configurazione tale che le condizioni seguenti sono rispettate: for all day $d$ and coin $c$, we have
	\[
		\sum_{u\in \{X,Y,Z\}} p(u,c,d) = (\top, 1)
	\]
	where we denote as ``sum'' the logical conjunction in $\Omega_I$: this means that day by day, the \emph{global} existence of the group of coins constantly attains the maximum; it is the \emph{local} existence that lowers when the initial conglomerate of coins is partitioned. Moreover,
	\[
		\begin{cases}
			\sum_{c_X\in C_X} p(X,c_X,V) = (\top,1) \\
			\sum_{c_Y\in C_Y} p(Y,c_Y,G) = (\top,1) \\
			\sum_{c_Z\in C_Z} p(Z,c_Z,V) = (\top,1)
		\end{cases}
	\]
	In an admissible configuration the subsets $ C_X, C_Y, C_Z $ can only attain an existence $p(g,c,d) \lneq (\top,1)$; that is no coin completely exists \emph{locally}. But for an hypothetical external observer, capable of adding up the forces with which the various parts of $C$ exist, the coins \emph{globally} exist `` in some secret way, of understanding forbidden to men'' (or rather, to $ X, Y, Z $).
	\begin{center}
		\begin{figure}
			\begin{tikzpicture}[xscale=6, yscale=4]
				\coordinate (1) at (0,0);
				\foreach \i/\j in {1/W,2/Th,3/Fr}{
						\draw[gray!90] (\i/4,0) node[below] {\tiny \j} -- (\i/4,1);
					}
				\node[below,gray!90] at (0,0) {\tiny Tu};
				\node[below,gray!90] at (1,0) {\tiny S};
				\coordinate (1) at (0,0);
				\coordinate (2) at (1/2, 1/5);
				\coordinate (3) at (3/4,1);
				\draw[ultra thick,red] (1) -- (2) node[right] {\scriptsize $1/5$} -- (3);
				\draw[thin,yellow] (1) -- (2) -- (3);
				\draw[blue] (1) -- (1/2,1) node[above] {\scriptsize $p(Y, c_Y)=(\top,1)$} -- (3/4,1/4) node[right] {\scriptsize $1/4$};
				\node[fill=gray!60] (cloud) at (1/4,1/2) {\large\color{black} \faCloud};
				\draw[thick] (0,0) rectangle (1,1);
			\end{tikzpicture}
			\caption{A pictorial representation of the truth forces of coins in different days, a piecewise linear model. $X$ is red, $Z$ is yellow, $Y$ is blue. Time is considered as a continuum marked at weekdays.}
		\end{figure}
	\end{center}
\end{example}
\begin{remark}\label{di_navi_e_numeri}
  L'aritmetica di Tlon; proposizioni a forza additiva; parallelismi tra la nave di Teseo e le nove monete.
\end{remark}  
\begin{remark}[Continuity for a proposition]\label{continuiti}
	Let $p : U \to \Omega_I$ be a proposition; here we investigate what does it mean for $p$ to be (globally) continuous with respect to the Euclidean topology on $I=[0,1]$, in the assumption that its domain of definition $U$ is metrizable (this is true for example when $U$ is a subset of space-time). The condition is that
	\[ \forall \epsilon > 0,\,\exists \delta > 0 : |x-y|< \delta \To |px-py| < \epsilon \]
	Da questo segue immediatamente che quando $p$ è continua nel suo dominio, i valori di verità di $p$ in configurazioni ``vicine'' in un senso opportuno sono dati con forze allo stesso modo vicine (chiaramente questa è una descrizione spannometrica della nozione di continuità\dots).

	All elementary topology results apply to such a proposition: the set of forces with which $p$ is true or false is a connected subset of $\Omega_I$, compact if $U$ was compact.
\end{remark}
\begin{example}[Discontinuity, Lo zaffiro di Taprobana]\label{bli}
	la lotteria a Babilonia come in \cite{babil}: proposizioni $p : U \to \Omega_{[0,1]}$ possono essere fortemente discontinue nelle variabili/contesto da cui dipendono: tali proposizioni descrivono eventi apparentemente caotici, innescati come termine finale di una catena di eventi tra loro disconnessi e paradossali;\footnote{Assumendo una base reale per lo spazio dei parametri da cui $p$ dipende, la sua dipendenza continua è la proprietà enunciata in \autoref{continuiti}; ciò significa che eventi vicini --nello spazio o nella consequenzialità temporale-- non possono avere valori di verità diversi, e forze ``vicine'' in un senso opportuno.} un modello di questo unvierso si trova probabilmente nella Babilonia di \cite{babil}, e nei ``sorteggi impersonali, di proposito indefinito'' che caratterizzano l'operato della Compagnia, azioni apparentemente scorrelate tra loro (scagliare ``nelle acque dell'Eufrate uno zaffiro di Taprobana''; sciogliere ``dal tetto d'una torre [\dots\unkern] un uccello''; togliere (o aggiungere) ``un granello di rena ai grani innumerevoli della spiaggia''; queste azioni hanno ``conseguenze, a volte, tremende''.
\end{example}
\begin{example}[Continuity: a few birds, a horse]\label{blu}
	Per quanto riguarda le proposizioni che sono continue nelle proprie variabili, invece, esempio canonico sono le ``tigri di cristallo'' e le ``torri di sangue'' di Tl\"{o}n: oggetti ed entità usuali, diversi da quelli ``classici'' per un dettaglio solo (il colore, la consistenza, il materiale di cui sono composte): la loro esistenza è sfumata, forte meno del massimo, cosicché tigri di carne e una torre di pietra sono tali che $p=(\top,1)$, le loro controparti su Tl\"on esistono con meno forza.

	Un altro esempio illustre è fatto da ogetti la cui forza di esistenza dipende in maniera \emph{monotona} e continua dai loro parametri: per esempio una proposizione $p$ può essere tanto più vera quanta più gente la osserva, perché ``le cose, su Tlön, si duplicano; ma tendono anche a cancellarsi e a  perdere i dettagli quando la gente le dimentichi. È classico l'esempio di  un'antica soglia, che perdurò finchè un mendicante venne a visitarla, e che alla  morte di colui fu perduta di vista. Talvolta pochi uccelli, un cavallo, salvarono le  rovine di un anfiteatro. ''

	In questa situazione, poniamo ad esempio che la forza di esistenza di alcune rovine --modellate come è ingenuo fare, come un corpo rigido $R$ nello spazio, dipenda dal numero dei suoi osservatori:
	\[\textstyle p(R, n) = \big(\top, 1-\frac{1}{n}\big)\]
\end{example}
Succedono cose interessanti anche a cambiare topologia su $I$: per esempio, su $[0,1]$ possiamo mettere brutalmente la topologia discreta; in questo modo $I$ è l'unione disgiunta dei suoi punti $\{ \{t\} \mid t\in [0,1]\}$, e il classifo è l'unione disgiunta di $[0,1]$ copie di $\{0,1\}$. (See \autoref{fig:berkeley} for a picture.)
\begin{example}[La campagna incendiata] 
  Il Berkeley idealista degli infiniti istanti di tempo continuo, disconnessi e incomunicabili: $\Omega_I = \coprod_{t : [0,1]} \{ 0 < 1 \}$. E' evidente come questa particolare struttura logica influenzi il linguaggio piegandolo a diventare l'istantaneismo berkeleyano: i termini sono costruiti per accrezione istantanea, per somma disgiunta dei costituenti e delle loro proprietà: ``aereo-chiaro sopra scuro rotondo''; oggetti determinati dalla loro simultaneità, e non da una dipendenza logica: il significato si costruisce per accrezione di istanti simultanei, e non per sequenzialità temporale (vedi \autoref{di_navi_e_numeri} per un legame tra questo principio, la natura additiva della forza di verità in Tlon, e la particolare forma dell'aritmetica Tloniana). 
  
  Il rifiuto della consequenzialità temporale per gli abitanti di Tlon sta nel passo 
  \begin{quote}
    \hspace{.5em} Spinoza attribuisce alla sua inesauribile divinità i modi del pensiero e dell'estensione; su Tlön, nessuno comprenderebbe la giustapposizione del  secondo (che caratterizza solo alcuni stati) e del primo, che è un sinonimo  perfetto del cosmo. In altre parole: non concepiscono che lo spaziale perduri  nel tempo. La percezione di una fumata all'orizzonte, e poi della campagna  incendiata, e poi della sigaretta mal spenta che provocò l'incendio, è  considerata un esempio di associazione di idee.
  \end{quote} 
  Ciò si lega anche al passo ``L'universo è paragonabile a quelle crittografie in cui non tutti i segni hanno un valore, e che solo è vero ciò che accade ogni trecento notti'': un mondo dove ogni trecento notti $p(x) =(\top,1)$, e per le successive 299 notti $p$ ha forza $<1$.
	\begin{center}
		\begin{figure}
			\begin{tikzpicture}[xscale=4, yscale=2]
				\fill[gray!30] (0,0) rectangle (1,1);
				\foreach \i in {.1,1,...,10}
				\draw[xshift=\i, yshift=-\i, ultra thin, fill=gray!30, opacity=.5] (0,0) rectangle (1,1);
				\draw[->, >=stealth, thin] (-.1,0) -- (-.1,1.5);
				\draw[->, >=stealth, thin] (-.1,0) -- (.5,-3/5);
				\draw[gray!60, xshift=.5cm, yshift=.25cm] (0,0) .. controls (.5,0) and (0,-.5) .. (.5,0);
			\end{tikzpicture}
      \caption{Il tempo come sequenza infinita, e infitamente suddivisibile, di istanti distinti: il paradosso Berkeleyano.}
      \label{fig:berkeley}
		\end{figure}
	\end{center}
\end{example}
\begin{example}

\end{example}
